---
title: "MA22015 2026 - Coursework"
output:
  html_document:
    df_print: paged
    toc: true
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

<!-- Anything written like this is an instruction about formatting the coursework. You should remove these before submission so they don't get included in the word count -->

```{=html}
<!-- Do not include your name in the output. Moodle will associate
your name with your upload. It does not matter what name you give to this file. Do not remove the table of contents (toc) -->
```

```{=html}
<!-- You may add cosmetic options only to this chunk. All other R chunks should be 
visible in your output -->
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment=NA)
```

<!-- Do not change the section titles for the questions. Keep an empty line above and below each section title to ensure they render correctly. -->

## Generative AI statement

<!-- Edit this R chunk -->

```{r}
# Type B: Generative AI is permitted as an assistive tool within the assessment.
# Describe how you used as AI here
# Write this as a code comment as I have done here so that this will not be included
# in the word count. Do not use code comments elsewhere as a means to include 
# additional content.
```

## Setup

<!-- Load in the data and any R packages you use here. -->

<!-- The R statement reading in your data should work for data placed in the same folder or directory as your coursework file. This is necessary so that I can run your R markdown file myself without editing. -->

```{r}
library(ggplot2)
polpres <- read.csv("polpres.csv")
View(polpres)
```

The wordcount:

<!-- Put the name of your Rmarkdown file within the quotes -->

```{r}
library(rmdwc)
rmdcount("cw26.Rmd")$words
```

## 1. Additional Voters

```{r}

candidates <- c("Bartoszewicz","Biejat","Braun","Hołownia","Jakubiak",
"Maciak","Mentzen","Nawrocki","Senyszyn","Stanowski",
"Trzaskowski","Woch","Zandberg")

polpres$round1 <- rowSums(polpres[, candidates])
polpres$round2 <- polpres$N2 + polpres$T2 + polpres$spoilt
polpres$additional <- polpres$round2 - polpres$round1

pp <- polpres[polpres$province != "", ]

lmod_before <- lm(additional ~ round1, data = pp)

par(mfrow = c(2, 2)) 
plot(lmod_before) 


lmod <- lm(log(additional) ~ log(round1), data = pp)

stud <- rstudent(lmod) 
pp$stud <- stud

n <- nrow(pp) 
threshold <- qt(0.05 / (2 * n), df = n - 2 - 1)
pp$outlier <- abs(pp$stud) > abs(threshold)

ggplot(pp, aes(log(round1), log(additional))) + geom_point() +
geom_smooth(method = "lm") + geom_text(data = pp[pp$outlier, ],
aes(label = county), hjust = -0.1, size = 3) + xlab("Log first round total votes") + ylab("Log additional voters in second round")

#do not include these - just for doing the coursework? 
names(stud) <- pp$county 
outliers <- stud[abs(stud) > abs(threshold)] 
outliers
par(mfrow = c(2, 2)) 
plot(lmod) 
# do not include these - just for doing the coureswork?

```

The counties 'zagranica' and 'statki' were excluded from this analysis
as they are not polish counties, and instead represent overseas
citizens. The scatter plot shows a strong positive linear relationship
between log(additonal) and log(round1). I fitted a linear model of these
two variables to identify counties with unusual changes relative to
their size. A log transformation was used after the original linear
model's diagonistic plots showed heteroscedasticity, with larger
counties having much higher variation. After applying this
transformation, the diagonistics were checked and showed no major
concerns. To identify outliers, studentized residuals were tested
against a Boferroni-corrected threshold, which produced three outliers:
gryficki (4.45), kołobrzeski (4.43), and kamieński (4.09). All 3 of
these counties are found in the zachodniopomorskie province as well as
being coastal towns, suggesting inflated voter counts reflect
holidaymakers voting there rather than at home.

## 2. Exit Poll

```{r}
p <- sum(polpres$T2) / (sum(polpres$T2) + sum(polpres$N2)) 
prob <- 1 - pbinom(500, 1000, p) 
print(p) 
print(prob)
```

Trzaskowski's national vote share (excluding spoilt ballots) is
`r round(p, 3)`. If we suppose 1000 randomly selected voters were
surveyed, the probability that a majority favour Trzaskowski is
`r round(prob, 3).` This figure seems likely and unsurprising given his
true share of just under 50%.

## 3. Turnout

```{r}
pp <- polpres[!is.na(polpres$province) & polpres$province != "", ]
pp$turnout <- (pp$N2 + pp$T2 + pp$spoilt) / pp$eligible2 * 100

lmod3 <- lm(turnout ~ province, data = pp)
anova(lmod3)

ggplot(pp, aes(x = reorder(province, turnout, FUN = median), y = turnout)) +
  geom_boxplot() +
  coord_flip() +
  xlab("Province") +
  ylab("Turnout (%)")
  
prov_means <- aggregate(turnout ~ province, data = pp, FUN = mean)
prov_means[order(prov_means$turnout), ]

# do not include these - just doing for couresework?
par(mfrow=c(2,2)); plot(lmod3)
# do not include these - just doing for the coureswork?
```

Zagranica and statki were excluded as the question specifies only those
voting in Poland. I computed round 2 turnout with (N2 + T2 + spoilt) /
eligible2 x 100 per county. Diagnostics plots were checked and showed no
concerns and validated one-way ANOVA assumptions. The ANOVA showed
significant differences between provinces (F = 14.38, p \< 0.001).
Opolskie (`r round(...)`%) and warmińsko-mazurskie (`r round(...)`%) had
exceptionally low turnout, while there were no clear exceptionally high
turnout provinces. This was supported by the sorted boxplot, which shows
clear separation at the lower end.

## 4. Spoilt Ballots

```{r}
pp <- polpres
pp <- pp[pp$county != "statki", ]
pp$round1 <- rowSums(pp[, candidates])
pp$round2 <- pp$N2 + pp$T2 + pp$spoilt
pp$spoilt_rate <- pp$spoilt / pp$round2
pp$T_share <- pp$Trzaskowski / pp$round1
lmod4 <- lm(spoilt_rate ~ T_share, data = pp)
summary(lmod4)

par(mfrow = c(2, 2))
plot(lmod4)
```

After checking initial diagnostic plots, statki was excluded due to its
clear outlier behavior. Spoilt rate was modelled against Trzaskowski's
first-round vote share. From the model summary we see a significant
positive coefficient, meaning that counties with stronger first-round
Trzaskowski support had higher spoilt rates. This suggests spoilt ballot
voters were more likely to be potential Trzaskowski supporters.

## 5. USA-style election

```{r}
pp <- polpres[!is.na(polpres$province) & polpres$province != "", ]

prov <- aggregate(cbind(T2, N2, eligible2) ~ province, data = pp, FUN= sum) 
prov$winner <- ifelse(prov$T2 >
prov$N2, "Trzaskowski", "Nawrocki")
prov$ec_votes <- prov$eligible2
prov$margin <- abs(prov$T2 - prov$N2) 

ec_T <- sum(prov$ec_votes[prov$winner == "Trzaskowski"]) 
ec_N <- sum(prov$ec_votes[prov$winner == "Nawrocki"]) 
ec_winner <- ifelse(ec_T > ec_N, "Trzaskowski", "Nawrocki") 

ec_winner_provs <- prov[prov$winner == ec_winner, ]
ec_winner_provs <- ec_winner_provs[order(ec_winner_provs$margin), ]

ec_lead <- abs(ec_T - ec_N) 
best <- Inf
best_combo <- NULL
for (k in 1:nrow(ec_winner_provs)) {
  combos <- combn(1:nrow(ec_winner_provs), k, simplify = FALSE)
  for (combo in combos) {
    ec_swing <- sum(ec_winner_provs$ec_votes[combo]) * 2
    if (ec_swing > ec_lead) {
      cost <- sum(floor(ec_winner_provs$margin[combo] / 2) + 1)
      if (cost < best) {
        best <- cost
        best_combo <- combo
      }
    }
  }
}

national_T <- sum(polpres$T2)
national_N <- sum(polpres$N2) 
national_margin <- abs(national_T - national_N) 
national_flips <- floor(national_margin / 2) + 1
```

Trzaskowski would win the election under EC rules (19.2M v 9.4M
electoral votes), despite losing to Nawrocki under the actual Polish
voting system. However, only 20,333 voter changes would be needed to
flip the electoral college outcome, compared to 184,796 changes for the
actual Polish result. This shows that the electoral system is much more
sensitive to small vote changes, with almost 100 times fewer switches
needed when comparing the two systems.

## 6. Transfer of first round votes

```{pp <- polpres}
pp$round1 <- rowSums(pp[, candidates])
pp$round2 <- pp$N2 + pp$T2 + pp$spoilt
pp$others <- pp$Bartoszewicz + pp$Jakubiak + pp$Maciak + 
             pp$Senyszyn + pp$Stanowski + pp$Woch
pp$latejoins <- pp$round2 - pp$round1

lmod6 <- lm(T2 ~ Biejat + Zandberg + Hołownia + Braun + Mentzen + 
             Trzaskowski + Nawrocki + others + latejoins - 1, data = pp)
summary(lmod6)
```
