---
title: "MA22015 2026 - Coursework"
output:
  html_document:
    df_print: paged
    toc: true
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment=NA)
```

## Generative AI statement

```{r}
# GenAI Usage:
# 1) Identifying political stances of candidates, to group into left/centre/right in Q4,
# 2) R code specificially for making ggplots,
# 3) Checking for spelling and grammatical inconsistencies.   
```

## Setup

```{r}
library(ggplot2)
polpres <- read.csv("polpres.csv")
```

The wordcount:

```{r}
library(rmdwc)
rmdcount("cw26.Rmd")$words
```

## 1. Additional Voters

```{r}

candidates <- c("Bartoszewicz","Biejat","Braun","Hołownia","Jakubiak",
"Maciak","Mentzen","Nawrocki","Senyszyn","Stanowski",
"Trzaskowski","Woch","Zandberg")

polpres$round1 <- rowSums(polpres[, candidates])
polpres$round2 <- polpres$N2 + polpres$T2 + polpres$spoilt
polpres$additional <- polpres$round2 - polpres$round1

pp <- polpres[polpres$province != "", ]
```

Zagranica' (foreign countries) and 'statki' (ships) have no province and
are not Polish counties, so were excluded. Additional voters were
calculated as the difference between round 2 total votes (N2 + T2 +
spoilt) and round 1 total votes across all 13 candidates.

```{r}

ggplot(pp, aes(round1, additional)) + geom_point() + 
  geom_smooth() + 
  xlab("First round total votes") + ylab("Additional voters in second round")
```

```{r}
lmod_before <- lm(additional ~ round1, data = pp)
par(mfrow = c(2, 2)) 
plot(lmod_before)
```

A linear model was initially fitted and round1 was a statistically
significant predictor of additional voters. However, diagnostic plots
show clear heteroscedasticity, with larger counties having much higher
variation. Observation 179 (Warsaw) has extremely high leverage due to
its large population and thus model influence, though it follows the
overall trend and is not an outlier.

```{r}
lmod_after <- lm(log(additional) ~ log(round1), data = pp)
plot(lmod_after, 1)

```

A log-log transformation was applied to both variables. This stabilises
the variance and produces interpretable model coefficients. Inspecting
the models diagnostics plots, no major concerns were seen and thus a
linear model appears to be a good fit. The model achieves an R\^2 of
0.8276, with log(round1) a statistically significant predictor with an
estimated coefficient of 0.89; implying that additional voters scales
approximately proportionally with county size.

```{r}
stud <- rstudent(lmod_after) 
pp$stud <- stud

n <- nrow(pp) 
threshold <- qt(0.05 / (2 * n), df = n - 2 - 1)
pp$outlier <- abs(pp$stud) > abs(threshold)

ggplot(pp, aes(log(round1), log(additional))) + geom_point() +
geom_smooth(method = "lm") + geom_text(data = pp[pp$outlier, ], 
aes(label = county), hjust = -0.1, size = 3) + xlab("Log first round total votes") + ylab("Log additional voters in second round")
```

To identify outliers, studentized residuals were tested against a
Bonferroni-corrected threshold, which produced three outliers: gryficki
(4.45), kołobrzeski (4.43), and kamieński (4.09). All 3 of these
counties are found in the zachodniopomorskie province as well as being
coastal towns, which may reflect holidaymakers voting there, though
other explanations are possible.

## 2. Exit Poll

```{r}
p <- sum(polpres$T2) / (sum(polpres$T2) + sum(polpres$N2)) 
```

Spoilt votes were excluded as these are unlikely to represent voters who
truthfully revealed their choice, and we cannot make any conclusions yet
about where these votes would have ended up.

Modelling the survey as 1000 independent draws from the voting
population, each voter's choice follows a Bernoulli trial, giving a
Binomial(1000, p) model where p is Trzaskowski's national vote share
(excluding spoilt ballots). This assumes voters are drawn independently
and uniformly at random, each truthfully revealing their choice with
constant probability p. The proportion of votes for 'Trzaskowski' was
0.49, marginally below Nawrocki's.

```{r}
prob <- 1 - pbinom(500, 1000, p) 
```

Using the binomial model, if we suppose 1000 randomly selected voters
were surveyed, the probability that a majority favour Trzaskowski is
0.28. This means there is roughly a 1 in 4 chance a random sample of
1000 voters would show a Trzaskowski majority, which is plausible given
his true share sits just below 50%. This means that the exit poll result
wasn't unlikely - it could easily happen by chance even when the true
share is below 50%.

## 3. Turnout

```{r}
pp <- polpres[!is.na(polpres$province) & polpres$province != "", ]
pp$turnout <- (pp$N2 + pp$T2 + pp$spoilt) / pp$eligible2 * 100
```

As before, zagranica and statki were excluded. Turnout was calculated as
(N2 + T2 + spoilt) / eligible2 × 100 for each county.

```{r}
lmod3 <- lm(turnout ~ province, data = pp)
anova(lmod3)
```

County-level turnout varies within each province, so we treat it as a
random variable and use one-way ANOVA to test whether average turnout
differs across provinces. Diagnostic plots were checked and the
residual-fitted and residuals-leverage plots showed no concerns. The Q-Q
plot showed minor departure from normality in the lower tail, but with
approximately 24 counties per province the Central Limit Theorem ensures
the F-test remains reliable. The ANOVA showed highly significant
differences in turnout between provinces (F(15, 364) =
`` `r round(anova(lmod3)$"F value"[1], 2)` ``, p \< 0.001).

```{r}
ggplot(pp, aes(x = reorder(province, turnout, FUN = median), y = turnout)) +
  geom_boxplot() +
  coord_flip() +
  xlab("Province") +
  ylab("Turnout (%)")
```

The F-test conclusion is supported by the boxplot, which shows a mixed
spread of province turnouts and a high degree of within-province
variability, especially for 'mazowieckie' and 'pomorskie'. 'Opolskie'
and 'warmińsko-mazurskie' clearly separate at the lower end while the
remaining provinces form a more gradual rise in turnout. At the upper
end, no province appears exceptionally high or distinct from the general
trend, with 'mazowieckie' having the highest turnout percentage, a
likely ranking given it contains 'Warsaw', the county with the highest
official turnout seen in the dataset.

## 4. Spoilt Ballots

```{r}
pp4 <- polpres
pp4$round1 <- rowSums(pp4[, candidates])
pp4$round2 <- pp4$N2 + pp4$T2 + pp4$spoilt
pp4$T_share <- pp4$Trzaskowski / pp4$round1
pp4$spoilt_rate <- pp4$spoilt / pp4$round2

ggplot(pp4, aes(T_share, spoilt_rate)) + geom_point() + 
  geom_smooth() + 
  xlab("Trzaskowski first-round vote share") + ylab("Spoilt rate")
```

```{r}
lmod4 <- lm(spoilt_rate ~ T_share, data = pp4)
plot(lmod4, 4)

```

Spoilt rate was modelled against Trzaskowski's first-round vote share.
Despite the clear high point in the scatter plot, no outlier values were
removed as cook's distance was examined and no observations exceeded the
conventional 0.5 threshold. The model summary shows a significant
positive coefficient (`r round(coef(lmod4)[2], 4)`, p \< 0.001), meaning
counties with stronger first-round Trzaskowski support had higher spoilt
rates. This suggests spoilt ballots were more common in areas with
centre-left support. The R² of `round(summary(lmod4)$r.squared, 3)` is
modest, but the significant coefficient is sufficient to identify the
direction of the political association.

```{r}
pp4$left_share <- (pp4$Biejat + pp4$Zandberg) / pp4$round1
pp4$right_share <- (pp4$Braun + pp4$Mentzen + pp4$Nawrocki) / pp4$round1
pp4$centre_share <- (pp4$Hołownia + pp4$Trzaskowski) / pp4$round1

cor(pp4[, c("left_share", "right_share", "centre_share")])
```

To investigate further, first-round vote shares were computed for
left-wing (Biejat, Zandberg), right-wing (Braun, Mentzen, Nawrocki) and
centre (Hołownia, Trzaskowski) blocs. The remaining minor candidates
were excluded from the classification as they are difficult to classify
politically. The correlation matrix shows near-perfect collinearity
(right-centre: -0.99), so a combined model would produce unreliable
coefficients. This is especially important given the
interpretation-nature of this task.

```{r}
pp4_long <- data.frame(
  spoilt_rate = rep(pp4$spoilt_rate, 3),
  share = c(pp4$left_share, pp4$right_share, pp4$centre_share),
  bloc = rep(c("Left", "Right", "Centre"), each = nrow(pp4))
)

ggplot(pp4_long, aes(share, spoilt_rate, color = bloc)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(method = "lm") + 
  xlab("Vote share") + ylab("Spoilt rate")
```

The fitted line plot confirms a clear positive trend. This is consistent
with the suggestion that more spoilt ballot voters intended to vote for
Trzaskowski. The broader political pattern is illustrated by overlaying
fitted lines for left, right, and centre vote shares against spoilt
rate; left and centre shares show a positive association while
right-wing share shows a negative one. This suggests spoilt ballot
voters came from areas with centre-left political leanings, though as
this is county-level data we cannot be certain about the intentions of
individual voters, a clear example of the ecological fallacy.

## 5. USA-style election

```{r}
pp <- polpres[!is.na(polpres$province) & polpres$province != "", ]

prov <- aggregate(cbind(T2, N2, eligible2) ~ province, data = pp, FUN= sum) 
prov$winner <- ifelse(prov$T2 > prov$N2, "Trzaskowski", "Nawrocki")
prov$ec_votes <- prov$eligible2
```

Given this question concerns provinces, 'Zagranica' and 'statki' were
excluded as they are not true polish counties and thus do not fall into
any given province. For each province, we calculate the winning
candidate who claims all electoral votes for that region. We represent
each province's electors by their number of eligible voters, mimicking
the traditional, proportional representation assigned to states.

```{r}
prov$margin <- abs(prov$T2 - prov$N2) 

ec_T <- sum(prov$ec_votes[prov$winner == "Trzaskowski"]) 
ec_N <- sum(prov$ec_votes[prov$winner == "Nawrocki"]) 
ec_winner <- ifelse(ec_T > ec_N, "Trzaskowski", "Nawrocki") 

ec_winner_provs <- prov[prov$winner == ec_winner, ]
ec_winner_provs <- ec_winner_provs[order(ec_winner_provs$margin), ]

ec_lead <- abs(ec_T - ec_N) 
best <- Inf
best_combo <- NULL
for (k in 1:nrow(ec_winner_provs)) {
  combos <- combn(1:nrow(ec_winner_provs), k, simplify = FALSE)
  for (combo in combos) {
    ec_swing <- sum(ec_winner_provs$ec_votes[combo]) * 2
    if (ec_swing > ec_lead) {
      cost <- sum(floor(ec_winner_provs$margin[combo] / 2) + 1)
      if (cost < best) {
        best <- cost
        best_combo <- combo
      }
    }
  }
}
```

Under the electoral college system, Trzaskowski wins, a direct reversal
of the actual result. This highlights the influence of electoral systems
and how their mechanisms can bring about vastly different outcomes when
it comes to determining presidency.

The minimum flip cost is found by searching all subsets of the winner's
provinces for the cheapest combination whose electoral college swing
exceeds the winner's lead.

Under the revised election system, only 20,333 voter changes across
mazowieckie and warmińsko-mazurskie would reverse the result. Despite
the high vote margin between candidates, narrow margins in large
provinces can be flipped very cheaply, swinging enormous electoral
college weight and reversing the election result for minimal voting
cost.

```{r}
national_T <- sum(polpres$T2)
national_N <- sum(polpres$N2) 
national_margin <- abs(national_T - national_N) 
national_flips <- floor(national_margin / 2) + 1
```

Under the actual Polish system, 184,796 vote changes are needed to alter
the outcome, almost 10 times more than the Electoral College. This is
due to the electoral system's sensitivity to small vote changes and
marginal shifts in regions shaping national outcomes. This contrasts
Poland's direct popular vote system, which reflects overall opinions and
voting changes more proportionally.

## 6. Transfer of first round votes

```{r}
pp <- polpres
pp$round1 <- rowSums(pp[, candidates])
pp$round2 <- pp$N2 + pp$T2 + pp$spoilt
pp$others <- pp$Bartoszewicz + pp$Jakubiak + pp$Maciak + pp$Senyszyn + pp$Stanowski + pp$Woch
pp$latejoins <- pp$round2 - pp$round1

ggplot(pp, aes(Hołownia, T2)) + geom_point() + 
  geom_smooth() + 
  xlab("Hołownia first-round votes") + ylab("Trzaskowski second-round votes")

lmod6 <- lm(T2 ~ Trzaskowski + Nawrocki + Biejat + Zandberg + Hołownia + Braun + Mentzen + others + latejoins - 1, data = pp)
```

Every second-round voter either supported a first-round candidate or
joined late, so the predictors account for all round 2 voters. The
intercept is removed since zero first-round voters should produce zero
second-round votes. To account for the stability of the finalists' voter
bases, their first-round totals were included as baseline controls. This
allows the model to better isolate how the losing candidates' voters
redistributed their support.

```{r}
plot(lmod6, c(1,2))
```

Diagnostic plots were checked. The residual-fitted plot shows variance
increasing with fitted values, which is expected for count data where
larger counties produce larger residuals. The Q-Q plot shows heavy tails
with observation 181 particularly extreme. Refitting the model without
this observation changed the magnitude of some coefficients but the
overall pattern of transfers remained the same, so it was retained.
Since the goal is to interpret coefficients as transfer proportions, a
transformation of the response would not be appropriate as the
coefficients would no longer have this interpretation.

```{r}
coefs <- round(coef(lmod6), 3)
transfer <- data.frame(
  Group = c("Biejat", "Zandberg", "Hołownia", "Braun", "Mentzen", "Late joiners"),
  Trzaskowski = coefs[c("Biejat","Zandberg","Hołownia","Braun","Mentzen","latejoins")],
  Nawrocki = round(1 - coefs[c("Biejat","Zandberg","Hołownia","Braun","Mentzen","latejoins")], 3)
)
transfer

#coef(lm(T2 ~ Trzaskowski + Nawrocki + Biejat + Zandberg + Hołownia + Braun + Mentzen + others + latejoins - 1, 
#        data = pp[-181, ]))[c("Biejat", "Zandberg", "Hołownia", "Braun", "Mentzen", "latejoins")]
```

Each coefficient estimates the proportion of that group transferring to
Trzaskowski, with approximately (1 − coefficient) going to Nawrocki. As
spoilt ballots are a small fraction of the round 2 total votes, they are
absorbed into these proportions and have little effect on the overall
conclusions.

This is an ecological regression, with individual transfer behaviour
inferred from aggregate county-level data. The linear model does not
constrain coefficients to [0, 1], and the predictors sum to round 2
totals for each county, introducing near-collinearity. These model
constraints cause the Trzaskowski, Nawrocki, and Zandberg coefficients
to fall outside [0, 1]. Trzaskowski and Nawrocki are control variables,
not quantities of interest, and Zandberg's value (1.045) is close enough
to 1 to interpret as an approximately complete transfer. The four
remaining eliminated candidates and late joiners remain interpretable.

The model estimates suggest Biejat and Zandberg (left-wing) voters
transferred heavily to Trzaskowski, Hołownia (centrist) voters leaned
towards Trzaskowski, Braun and Mentzen (right-wing) voters went
predominantly to Nawrocki, and late joiners favoured Trzaskowski at
approximately 75%. These estimated transfers are consistent with
political positioninG.
