---
title: "MA22015 2026 - Coursework"
output:
  html_document:
    df_print: paged
    toc: true
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

<!-- Anything written like this is an instruction about formatting the coursework. You should remove these before submission so they don't get included in the word count -->

```{=html}
<!-- Do not include your name in the output. Moodle will associate
your name with your upload. It does not matter what name you give to this file. Do not remove the table of contents (toc) -->
```

```{=html}
<!-- You may add cosmetic options only to this chunk. All other R chunks should be 
visible in your output -->
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment=NA)
```

<!-- Do not change the section titles for the questions. Keep an empty line above and below each section title to ensure they render correctly. -->

## Generative AI statement

<!-- Edit this R chunk -->

```{r}
# Type B: Generative AI is permitted as an assistive tool within the assessment.
# Describe how you used as AI here
# Write this as a code comment as I have done here so that this will not be included
# in the word count. Do not use code comments elsewhere as a means to include 
# additional content.
```

## Setup

<!-- Load in the data and any R packages you use here. -->

<!-- The R statement reading in your data should work for data placed in the same folder or directory as your coursework file. This is necessary so that I can run your R markdown file myself without editing. -->

```{r}
library(ggplot2)
polpres <- read.csv("polpres.csv")
```

The wordcount:

<!-- Put the name of your Rmarkdown file within the quotes -->

```{r}
library(rmdwc)
rmdcount("cw26.Rmd")$words
```

## 1. Additional Voters

```{r}

candidates <- c("Bartoszewicz","Biejat","Braun","Hołownia","Jakubiak",
"Maciak","Mentzen","Nawrocki","Senyszyn","Stanowski",
"Trzaskowski","Woch","Zandberg")

polpres$round1 <- rowSums(polpres[, candidates])
polpres$round2 <- polpres$N2 + polpres$T2 + polpres$spoilt
polpres$additional <- polpres$round2 - polpres$round1

pp <- polpres[polpres$province != "", ]
```

From initial dataset analysis, the two counties 'zagranica' and 'statki'
have no listed province. From exploring this further, these both
translate to 'foreign countries' and 'ships'. In light of this, these
two were excluded from the dataset for this question as the focus on
counties implies this is a reasonable adjustment to make. Additional
voters were calculated by taking the difference of voters across all 13
candidates, and the number of spoilt, 'Nawrocki' and 'Trzaskowski' votes
from round 2.

```{r}
lmod_before <- lm(additional ~ round1, data = pp)
par(mfrow = c(2, 2)) 
plot(lmod_before)
```

Initially, a linear model of 'round1' against 'additional' was fitted,
which found round1 to be a statistically significant predictor variable
for additional voters. However, after checking diagnostics plots it
became obvious quickly that a linear model was not valid here. There are
clear signs of heteroscedasticity, with larger counties having much
higher variation. Additionally, the residuals v leverage plot reveals
that observation 179 has an extremely high leverage which suggests it
disproportionately influences the model fit.

```{r}
lmod_after <- lm(log(additional) ~ log(round1), data = pp)
plot(lmod_after, 1)

```

A log-log transformation was applied to both variables. This stabilises
the variance and produces interpretable model coefficients. Inspecting
the models diagnostics plots, no major concerns were seen and thus a
linear model appears to be a good fit. The model achieves an R\^2 of
0.8276, with log(round1) a statistically significant predictor with an
estimated coefficient of 0.89; implying that additional voters scales
approximately proportionally with county size.

```{r}
stud <- rstudent(lmod_after) 
pp$stud <- stud

n <- nrow(pp) 
threshold <- qt(0.05 / (2 * n), df = n - 2 - 1)
pp$outlier <- abs(pp$stud) > abs(threshold)

ggplot(pp, aes(log(round1), log(additional))) + geom_point() +
geom_smooth(method = "lm") + geom_text(data = pp[pp$outlier, ], 
aes(label = county), hjust = -0.1, size = 3) + xlab("Log first round total votes") + ylab("Log additional voters in second round")
```

To identify outliers, studentized residuals were tested against a
Bonferroni-corrected threshold, which produced three outliers: gryficki
(4.45), kołobrzeski (4.43), and kamieński (4.09). All 3 of these
counties are found in the zachodniopomorskie province as well as being
coastal towns, suggesting inflated voter counts reflect holidaymakers
voting there rather than at home.

## 2. Exit Poll

```{r}
p <- sum(polpres$T2) / (sum(polpres$T2) + sum(polpres$N2)) 
```

Spoilt votes were excluded as these are unlikely to represent voters who
truthfully revealed their choice, and we cannot make any conclusions yet
about where these votes would have ended up.

Modelling the survey as 1000 independent draws from the voting
population, each voter's choice follows a Bernoulli trial, giving a
Binomial(1000, p) model where p is Trzaskowski's national vote share
(excluding spoilt ballots). This assumes each of the 1000 voters is
drawn independently and uniformly at random from the national voting
population, and that each voter truthfully reveals their choice with a
constant probability p across the sample. In practice, exit polls often
use clustered sampling, which may violate the independence assumption,
but for this simplified model the binomial is appropriate. The
proportion of votes for 'Trzaskowski' across the second round of voting
was found to be 0.49, only marginally smaller than that of the eventual
winner 'Nawrocki'.

```{r}
prob <- 1 - pbinom(500, 1000, p) 
```

Using the binomial model, if we suppose 1000 randomly selected voters
were surveyed, the probability that a majority favour Trzaskowski is
0.28. This means there is roughly a 1 in 4 chance a random sample of
1000 voters would show a Trzaskowski majority, which is plausible given
his true share sits just below 50%. This means that the exit poll result
wasn't unlikely - it could easily happen by chance even when the true
share is below 50%.

## 3. Turnout

```{r}
pp <- polpres[!is.na(polpres$province) & polpres$province != "", ]
pp$turnout <- (pp$N2 + pp$T2 + pp$spoilt) / pp$eligible2 * 100
```

'Zagranica' and 'statki' were excluded here as they do not represent
voting in Poland. To find the turnout figures for round 2, I calculated
(N2 + T2 + spoilt) / eligible2 × 100 for each county, which gives a
standardised measure of each county's voting proportion.

```{r}
lmod3 <- lm(turnout ~ province, data = pp)
plot(lmod3)
summary(lmod3)
anova(lmod3)
```

County-level turnout varies within each province, so we treat it as a
random variable and use one-way ANOVA to test whether average turnout
differs across provinces. Diagnostic plots were checked and the
residual-fitted and residuals-leverage plots showed no concerns. The Q-Q
plot showed minor departure from normality in the lower tail, but with
approximately 24 counties per province the Central Limit Theorem ensures
the F-test remains reliable. The ANOVA showed highly significant
differences in turnout between provinces (F(15, 364) =
`` `r round(anova(lmod3)$"F value"[1], 2)` ``, p \< 0.001).

```{r}
ggplot(pp, aes(x = reorder(province, turnout, FUN = median), y = turnout)) +
  geom_boxplot() +
  coord_flip() +
  xlab("Province") +
  ylab("Turnout (%)")
```

The F-test conclusion is supported by the boxplot, which shows a mixed
spread of province turnouts and a high degree of within-province
variability, especially for 'mazowieckie' and 'pomorskie'. 'Opolskie'
and 'warmińsko-mazurskie' clearly separate at the lower end while the
remaining provinces form a more gradual rise in turnout. At the upper
end, no province appears exceptionally high or distinct from the general
trend, with 'mazowieckie' having the highest turnout percentage, a
likely ranking given it contains 'Warsaw', the county with the highest
official turnout seen in the dataset.

## 4. Spoilt Ballots

```{r}
pp$T_share <- pp$Trzaskowski / pp$round1
pp$spoilt_rate <- pp$spoilt / pp$round2

ggplot(pp, aes(T_share, spoilt_rate)) + geom_point() + 
  geom_smooth(method = "lm") + 
  xlab("Trzaskowski first-round vote share") + ylab("Spoilt rate")

lmod4 <- lm(spoilt_rate ~ T_share, data = pp)
plot(lmod4)
summary(lmod4)
```

Spoilt rate was modelled against Trzaskowski's first-round vote share.
No outlier values were removed as cook's distance was examined and no
observations exceeded the conventional 0.5 threshold. The model summary
shows a significant positive coefficient (`r round(coef(lmod4)[2], 4)`,
p \< 0.001), meaning counties with stronger first-round Trzaskowski
support had higher spoilt rates. This suggests spoilt ballots were more
common in areas with centre-left support. The R² of
`r round(summary(lmod4)$r.squared, 3)` is modest, but the significant
coefficient is sufficient to identify the direction of the political
association.

```{r}
pp$left_share <- (pp$Biejat + pp$Zandberg) / pp$round1
pp$right_share <- (pp$Braun + pp$Mentzen + pp$Nawrocki) / pp$round1
pp$centre_share <- (pp$Hołownia + pp$Trzaskowski) / pp$round1

cor(pp[, c("left_share", "right_share", "centre_share")])
```

Spoilt rate was calculated as the proportion of spoilt ballots in the
second round total. To investigate the political views of spoilt ballot
voters, first-round vote shares were computed for left-wing (Biejat,
Zandberg), right-wing (Braun, Mentzen, Nawrocki) and centre (Hołownia,
Trzaskowski) blocs. The remaining minor candidates were excluded from
the classification as they are difficult to classify politically without
detailed, contextual knowledge of their viewpoints.

The correlation matrix between these three shares shows near-perfect
collinearity (right-centre correlation of -0.99), so fitting a model
with all three as predictors would produce unreliable coefficients. This
is especially important given the interpretation-nature of this task.

```{r}
pp_long <- data.frame(
  spoilt_rate = rep(pp$spoilt_rate, 3),
  share = c(pp$left_share, pp$right_share, pp$centre_share),
  bloc = rep(c("Left", "Right", "Centre"), each = nrow(pp))
)

ggplot(pp_long, aes(share, spoilt_rate, color = bloc)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(method = "lm") + 
  xlab("Vote share") + ylab("Spoilt rate")
```

The fitted line plot confirms a clear positive trend. This is consistent
with the concern that more spoilt ballot voters intended to vote for
Trzaskowski. The broader political pattern is illustrated by overlaying
fitted lines for left, right, and centre vote shares against spoilt
rate; left and centre shares show a positive association while
right-wing share shows a negative one. This suggests spoilt ballot
voters came from areas with centre-left political leanings, though as
this is county-level data we cannot be certain about the intentions of
individual voters, a clear example of the ecological fallacy.

## 5. USA-style election

```{r}
pp <- polpres[!is.na(polpres$province) & polpres$province != "", ]

prov <- aggregate(cbind(T2, N2, eligible2) ~ province, data = pp, FUN= sum) 
prov$winner <- ifelse(prov$T2 > prov$N2, "Trzaskowski", "Nawrocki")
prov$ec_votes <- prov$eligible2
```

Given this question concerns provinces, 'Zagranica' and 'statki' were
excluded as they are not true polish counties and thus do not fall into
any given province. For each province, we calculate the winning
candidate who claims all electoral votes for that region. We represent
each province's electors by their number of eligible voters, mimicking
the traditional, proportional representation assigned to states.

```{r}
prov$margin <- abs(prov$T2 - prov$N2) 

ec_T <- sum(prov$ec_votes[prov$winner == "Trzaskowski"]) 
ec_N <- sum(prov$ec_votes[prov$winner == "Nawrocki"]) 
ec_winner <- ifelse(ec_T > ec_N, "Trzaskowski", "Nawrocki") 

ec_winner_provs <- prov[prov$winner == ec_winner, ]
ec_winner_provs <- ec_winner_provs[order(ec_winner_provs$margin), ]

ec_lead <- abs(ec_T - ec_N) 
best <- Inf
best_combo <- NULL
for (k in 1:nrow(ec_winner_provs)) {
  combos <- combn(1:nrow(ec_winner_provs), k, simplify = FALSE)
  for (combo in combos) {
    ec_swing <- sum(ec_winner_provs$ec_votes[combo]) * 2
    if (ec_swing > ec_lead) {
      cost <- sum(floor(ec_winner_provs$margin[combo] / 2) + 1)
      if (cost < best) {
        best <- cost
        best_combo <- combo
      }
    }
  }
}
```

Trzaskowski would in fact have won the 2025 Polish election, had the
electoral college system been used. This is a direct reversal of
Nawrocki being the actual winner, highlighting the influence of
electoral systems and how their mechanisms can bring about vastly
different outcomes when it comes to determining presidency.

The minimum flip cost is found by searching all subsets of the winner's
provinces for the cheapest combination whose electoral college swing
exceeds the winner's lead.

Under the revised election system, only 20,333 voter changes would be
needed to reverse the election outcome, across two provinces
'mazowieckie' and 'warmińsko-mazurskie'. Despite the high vote margin
between candidates, narrow margins in large provinces can be flipped
very cheaply, swinging enormous electoral college weight and reversing
the election result for minimal voting cost.

```{r}
national_T <- sum(polpres$T2)
national_N <- sum(polpres$N2) 
national_margin <- abs(national_T - national_N) 
national_flips <- floor(national_margin / 2) + 1
```

The actual number of votes needed to alter the outcome of the election
under the Polish electoral system is 184,796, which represents the raw
difference in candidate votes between the two options of round 2.
Clearly, this number is almost 10 times larger than that of the
electoral college system. This is due to the electoral system's
sensitivity to small vote changes and marginal shifts in regions shaping
national outcomes. This contrasts Poland's direct popular vote system,
which reflects overall opinions and voting changes more proportionally.

## 6. Transfer of first round votes

```{r}
pp <- polpres
pp$round1 <- rowSums(pp[, candidates])
pp$round2 <- pp$N2 + pp$T2 + pp$spoilt
pp$others <- pp$Bartoszewicz + pp$Jakubiak + pp$Maciak + pp$Senyszyn + pp$Stanowski + pp$Woch
pp$latejoins <- pp$round2 - pp$round1
lmod6 <- lm(T2 ~ Trzaskowski + Nawrocki + Biejat + Zandberg + Hołownia + Braun + Mentzen + others + latejoins - 1, data = pp)
```

Every second-round voter either supported a first-round candidate or
joined late, so the predictors account for all round 2 voters. The
intercept is removed since zero first-round voters should produce zero
second-round votes. To account for the stability of the finalists' voter
bases, their first-round totals were included as baseline controls. This
allows the model to better isolate how the losing candidates' voters
redistributed their support.

```{r}
plot(lmod6)
```

Diagnostic plots were checked. The residual-fitted plot shows variance
increasing with fitted values, which is expected for count data where
larger counties produce larger residuals. The Q-Q plot shows heavy tails
with observations 179 and 181 particularly extreme, as well as both
points having high leverage. Refitting the model without these
observations changed the magnitude of some coefficients but the overall
pattern of transfers remained the same, so both were retained. Since the
goal is to interpret coefficients as transfer proportions, a
transformation of the response would not be appropriate as the
coefficients would no longer have this interpretation.

```{r}
coefs <- round(coef(lmod6), 3)
transfer <- data.frame(
  Group = c("Biejat", "Zandberg", "Hołownia", "Braun", "Mentzen", "Late joiners"),
  Trzaskowski = coefs[c("Biejat","Zandberg","Hołownia","Braun","Mentzen","latejoins")],
  Nawrocki = round(1 - coefs[c("Biejat","Zandberg","Hołownia","Braun","Mentzen","latejoins")], 3)
)
transfer

coef(lm(T2 ~ Trzaskowski + Nawrocki + Biejat + Zandberg + Hołownia + Braun + Mentzen + others + latejoins - 1, 
        data = pp[-c(179, 181), ]))[c("Biejat", "Zandberg", "Hołownia", "Braun", "Mentzen", "latejoins")]
```

Each coefficient estimates the proportion of that group transferring to
Trzaskowski, with approximately (1 − coefficient) going to Nawrocki. As
spoilt ballots are a small fraction of the round 2 total votes, they are
absorbed into these proportions and have little effect on the overall
conclusions.

This is an ecological regression, with individual transfer behaviour
inferred from aggregate county-level data. The linear model does not
constrain coefficients to [0, 1], and the predictors sum to round 2
totals for each county, introducing near-collinearity. These model
constraints cause the Trzaskowski, Nawrocki, and Zandberg coefficients
to fall outside [0, 1]. Trzaskowski and Nawrocki are control variables,
not quantities of interest, and Zandberg's value (1.045) is close enough
to 1 to interpret as an approximately complete transfer. The four
remaining eliminated candidates and late joiners remain interpretable.

The overall pattern is clear: Biejat and Zandberg (left-wing) voters
transferred heavily to Trzaskowski, Hołownia (centrist) voters leaned
towards Trzaskowski, Braun and Mentzen (right-wing) voters went
predominantly to Nawrocki, and late joiners favoured Trzaskowski at
roughly 75%. These transfers align with political positioning, with
left-wing and centrist voters supporting the more liberal Trzaskowski
while right-wing voters backing the conservative Nawrocki.
